PyPy: raising the level for Dynamic Language implementations

### 2007-05-27 talk/dyla2007/dyla.tex
try to steal the good ideas of holgers abstract

Dynamic languages are usually implemented in lower level environments, 
traditionally in C/Posix.  Recently .NET and the JVM are emerging as 
new higher-level environments, offering more built-in support for 
memory management, threading models, dynamic optimization techniques
and also offering an abundance of libraries for application development.   
However, manually implementing languages for a standardized runtime environment
may not be the last word on bringing dynamic language implementations to a higher level. 
The PyPy project successfully experimented with an approach that 
uses extensive meta-programming techniques to analyse and transform the 
single source of an language "specification" and automatically adds
lower level and runtime-dependent aspects, successfully producing 
efficent virtual machines for an runtime environment of choice. 
The meta-programming facilities are largely re-useable for implementing 
different computer languages. We thus argue in this paper that 
one should not write interpreters manually and dependent on lower 
level and runtime details - but rather explore and exploit 
meta-programming facilities which themselves can be written 
and tested in a very high language, overall raising the level 
of implementing dynamic languages.  While there is considerable 
complexity liying in such a meta-programming tool chain, it is 
re-usable for different language implementations and generally
helps to avoid having to implement language semantics and 
optimizations redundantly.

\footnote{This research is partially supported by the EU funded %research
 project:
IST 004779 PyPy (PyPy: Implementing Python in Python).} \\

>>>>

Dynamic languages are traditionally implemented by writing a virtual machine for
them, centered around an interpreter and/or a built-in compiler and providing
the object model and memory management. When a language becomes more successful,
the limitations of such an implementation lead to the emergence of alternative
implementations that try to solve some of the problems. Another reason for new
implementations is the desire to have the language integrate well with existing,
well-tuned object-oriented virtual machine like Java's. In this paper, we
describe the mechanisms that lead to an abundance of implementations, and
explore some of the limitations of standard VMs.  We propose a different
complementary alternative to writing VMs by hand and dealing with low-level
details: flexibly generating virtual machines from a single abstract language
``specification''.

This approach has been validated by the PyPy project that tries to explore
meta-programming facilities with which we can automatically insert many features
and low-level details into an automatically generated virtual machines --
including good just-in-time compilers tuned to the dynamic language at hand.
We thus argue in this paper that one should not write interpreters manually but
rather use meta-programming techniques and raise the overall level at which
dynamic languages are implemented. We believe this to be ultimately a better
investment of efforts than the development of more advanced general-purpose OO
VMs. \footnote{This research is partially supported by the EU funded
 project: IST 004779 PyPy (PyPy: Implementing Python in Python).} \\
